RProtoBuf: Efficient Cross-Language Data Serialization in R

### 2014-01-07 papers/rjournal/eddelbuettel-stokely.Rnw
add section on opencpu
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@724 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% !TeX root = RJwrapper.tex
% We don't want a left margin for Sinput or Soutput for our table 1.
%\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=0em}
%\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=0em}
%\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}
% Setting the topsep to 0 reduces spacing from input to output and
% improves table 1.
\fvset{listparameters={\setlength{\topsep}{0pt}}}
\renewenvironment{Schunk}{\vspace{\topsep}}{\vspace{\topsep}}

@@

>>>>

% We don't want a left margin for Sinput or Soutput for our table 1.
%\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=0em}
%\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=0em}
%\DefineVerbatimEnvironment{Scode}{Verbatim}{xleftmargin=2em}
% Setting the topsep to 0 reduces spacing from input to output and
% improves table 1.
\fvset{listparameters={\setlength{\topsep}{0pt}}}
\renewenvironment{Schunk}{\vspace{\topsep}}{\vspace{\topsep}}

@@

\section{Application: Protocol Buffers for Data Interchange in Web Services}

As the name implies, the primary application of protocol buffers is
data interchange in the context of inter-system communications. 
Network protocols such as HTTP describe procedures on client-server
communication, i.e. how to iniate requests, authenticate, send messages, 
etc. However, network 
protocols generally do not regulate \emph{content} of messages: they allow
transfer of any media type, such as web pages, files or video.
When designing systems where various components require exchange of specific data
structures, we need something on top of the protocol that prescribes 
how these structures are to be respresented in messages (buffers) on the
network. Protocol buffers solve exactly this problem by providing
a cross platform method for serializing arbitrary structures into well defined
messages, that can be exchanged using any protocol. The descriptors
(\texttt{.proto} files) are used to formally define the interface of a
remote API or network application. Libraries to parse and generate protobuf
messages are available for many programming languages, making it 
relatively straight forward to implement clients and servers.

\subsection{Interacting with R through HTTPS and Protocol Buffers}

One example of a system that supports protocol buffers to interact
with R is OpenCPU \citep{opencpu}. OpenCPU is a framework for embedded statistical 
computation and reproducible research based on R and Latex. It exposes a 
HTTP(S) API to access and manipulate R objects and allows for performing 
remote R function calls. Clients do not need to understand 
or generate any R code: HTTP requests are automatically mapped to 
function calls, and arguments/return values can be posted/retrieved
using several data interchange formats, such as protocol buffers.  
OpenCPU uses the \texttt{serialize\_pb} and \texttt{unserialize\_pb} functions
from the \texttt{RProtoBuf} package to convert between R objects and protobuf
messages. Therefore, clients need the \texttt{rexp.proto} descriptor mentioned
earlier to parse and generate protobuf messages when interacting with OpenCPU.

\subsection{HTTP GET: Retrieving an R object}

The \texttt{HTTP GET} method is used to read a resource from OpenCPU. For example,
to access the dataset \texttt{Animals} from the package \texttt{MASS}, a 
client performs the following HTTP request:

\begin{verbatim}
  GET https://public.opencpu.org/ocpu/library/MASS/data/Animals/pb
\end{verbatim}
The postfix \texttt{/pb} in the URL tells the server to send this
object in the form of a protobuf message. Alternative formats include 
\texttt{/json}, \texttt{/csv}, \texttt{/rds} and others. If the request
is successful, OpenCPU returns the serialized object with HTTP status 
code 200 and HTTP response header \texttt{Content-Type: application/x-protobuf}. 
The latter is the conventional MIME type that formally notifies the client to
interpret the response as a protobuf message.

Because both HTTP and Protocol Buffers have libraries available for many 
languages, clients can be implemented in just a few lines of code. Below
example code for both R and Python that retrieve a dataset from R with 
OpenCPU using a protobuf message. In R, we use the HTTP client from 
the \texttt{httr} package \citep{httr}, and the protobuf
parser from the \texttt{RProtoBuf} package. In this illustrative example we
download a dataset which is part of the base R distribution, so we can actually
verify that the object was transferred without loss of information.

<<eval=FALSE>>=
# Load packages
library(RProtoBuf)
library(httr)

# Retrieve and parse message
req <- GET ('https://public.opencpu.org/ocpu/library/MASS/data/Animals/pb')
output <- unserialize_pb(req$content)

# Check that no information was lost
identical(output, MASS::Animals)
@
This code suggests a method for exchanging objects between R servers, however this can 
also be done without protocol buffers. The main advantage of using an inter-operable format 
is that we can actually access R objects from within another
programming language. For example, in a very similar fasion we can retrieve the same
dataset in a Python client. To parse messages in Python, we first compile the 
\texttt{rexp.proto} descriptor into a python module using the \texttt{protoc} compiler:

\begin{verbatim}
  protoc rexp.proto --python_out=.
\end{verbatim}
This generates python module called \texttt{rexp\_pb2.py}, containing both the 
descriptor information as well as methods to read and manipulate the R object 
message. In the example below we use the HTTP client from the \texttt{urllib2}
module.

\begin{verbatim}
# Import modules
import urllib2
from rexp_pb2 import REXP

# Retrieve message
req = urllib2.Request('https://public.opencpu.org/ocpu/library/MASS/data/Animals/pb')
res = urllib2.urlopen(req)

# Parse rexp.proto message
msg = REXP()
msg.ParseFromString(res.read())
print(msg)
\end{verbatim}
The \texttt{msg} object contains all data from the Animals dataset. From here we
can easily extract the desired fields for further use in Python.

\subsection{HTTP POST: Calling an R function}

The example above shows how the \texttt{HTTP GET} method retrieves a 
resource from OpenCPU, for example an R object. The \texttt{HTTP POST} 
method on the other hand is used for calling functions and running scripts, 
which is the primary purpose of the framework. As before, the \texttt{/pb} 
postfix requests to retrieve the output as a protobuf message, in this
case the function return value. However, OpenCPU allows us to supply the
arguments of the function call in the form of protobuf messages as well.
This is a bit more work, because clients needs to both generate messages 
containing R objects to post to the server, as well as retrieve and parse
protobuf messages returned by the server. Using protocol buffers to post
function arguments is not required, and for simple (scalar) arguments 
the standard \texttt{appliation/www-url-encoded} format might be sufficient.
However, with protocol buffers the client can perform function calls with
more complex arguments such as R vectors or lists. The result is a complete
RPC system to do arbitrary R function calls from within 
any programming language.

The following example R client code performs the remote function call 
\texttt{stats::rnorm(n=42, mean=100)}. The function arguments (in this
case \texttt{n} and \texttt{mean}) as well as the return value (a vector
with 42 random numbers) are transferred using a protobuf message. RPC in
OpenCPU works like the \texttt{do.call} function in R, hence all arguments
are contained within a list.

<<>>=
#requires httr >= 0.2.99
library(httr)
library(RProtoBuf)

args <- list(n=42, mean=100)
payload <- serialize_pb(args, NULL)

req <- POST (
  url = "https://public.opencpu.org/ocpu/library/stats/R/rnorm/pb",
  body = payload,
  add_headers (
    "Content-Type" = "application/x-protobuf"
  )
)

#This is the output of stats::rnorm(n=42, mean=100)
output <- unserialize_pb(req$content)
print(output)
@
The OpenCPU server basically performs the following steps to process the above RPC request:

<<eval=FALSE>>=
fnargs <- unserialize_pb(inputmsg)
val <- do.call(stats::rnorm, fnargs)
outputmsg <- serialize_pb(val)
@
In reality the OpenCPU provides a lot of meta functionality such as handling
of sessions, exceptions, security, and much more. OpenCPU also makes it possible to store
output of a function call on the server, instead of directly retrieving it. Thereby 
objects can be shared with other users or used as arguments in a subsequent
function call. But in its essence, the HTTP API provides a simple way to perform remote 
R function calls over HTTPS. The same request can be performed in Python as follows:

\begin{verbatim}
import urllib2;
from rexp_pb2 import *;

#create the post payload, i.e. list(n=42, mean=100)
payload = REXP(
  rclass = 5,
    rexpValue = [
      REXP(rclass = 2, realValue = [42]), 
      REXP(rclass = 2, realValue = [100])
    ],
    attrName = [
      "names"
    ],
    attrValue = [
      REXP(rclass = 0, stringValue = [STRING(strval="n"), STRING(strval="mean")])
    ]
);

#HTTP POST
req = urllib2.Request(
  "https://public.opencpu.org/ocpu/library/stats/R/rnorm/pb", 
  data = payload.SerializeToString(), 
  headers = {
    'Content-type': 'application/x-protobuf'
  }
)
res = urllib2.urlopen(req);

#parse output pb
msg = REXP();
msg.ParseFromString(res.read());

#the return value is a double vector in this case
print(msg.realValue);
\end{verbatim}
